
```{r}
#| include: false
library(dsld)
library(qeML)
```

{{< pagebreak >}}

# Linear Statistical Models

This chapter could have been titled, say, "Optimization, Part I," since
many applications of linear algebra involve minimization or maximization
of some kind, and this chapter will involve calculus derivatives.  But
the statistical applications of linear algebra are equally important.

## Linear Regression through the Origin

Let's consider the **Nile**  dataset built-in to R. It is a time series,
one measurement per year.

```{r}
head(Nile)
# predict current year from previous year?
n1 <- Nile[-(length(Nile))]
head(n1)
n2 <- Nile[-1]
head(n2)
plot(n1,n2,cex=0.4,xlim=c(0,1400),yli=c(0,1400))
```

We would like to fit a straight line through that data point cloud. We
might have two motivations for doing this:

* The line might serve as nice summary of the data.

* More formally, let $C$ and $V$ denote the current and previous year's
measurements..Then the model 

$$E(C | V) = \beta V$$ 

may be useful. Here the slope $\beta$ is an unknown value to be
estimated from the data.  [Readers with some background in linear
regression models should note that this assumption of a linear trend
through the origin is the only assumption we are making here. Nothing on
normal distributions etc.]{.column-margin}

::: {.callout-important}
## Model validity
The great statistician George Box once said, "All models are wrong but
some are useful." All data scientists should keep this at the forefronts
of their minds.
:::

### Least squares approach
[This too is something all data
scientists should keep at the forefronts of their minds. We are always
working with sample data, subject to intersample variations. The
quantity $b$ is a random variable.]{.column-margin}
Let $b$ denote our estimate. How should we obtain it?
We wish to estimate $\beta$ from our data, which we regard as a sample
from the data generating process.
Denote our data by $(C_i,V_i), i = 1,...,100$. 

Pretend for a moment that we don't know, say, $C_{28}$. Using our
estimated $\beta$, our predicted value would be $b V_{28}$.  Our squared
prediction error would then be $(C_{28} - b W_{28})^2$.

Well, we actually do know $C_{28}$ (and the others in our data), so we
can answer the question:

<blockquote>

In our search for a good value of $b$, we can ask how well we would
predict our known data, using that candidate value of $b$ in our data.
Our total squared prediction error would be

$$
\sum_{i=1}^{100} [C_{i} - b V_{i} )^2
$$

A natural choice for $b$ would be the value that minimizes this
quantity.
[Why not look at the absolute value instead of
the square? The latter makes the math flow well, as will be seen
shortly.]{.column-margin}

</blockquote>

### Calculation

As noted, our choice for $b$ will be the minimizer of

$$
\sum_{i=1}^{100} (C_{i} - b V_{i})^2
$$

This is a straightforward calculus problem. Setting

$$
0 = \frac{d}{db} \sum_{i=1}^{100} (C_{i} - b V_{i} )^2 =
-2 \sum_{i=1}^{100} (C_{i} - b V_{i}) V_i
$$

and solving $b$, we find that

$$
b = \frac{\sum_{i=1}^n C_i V_i}{\sum_{i=1}^nV_i^2}
$$

### R code

```{r}
lm(n2 ~ n1-1)
```



This says, "Fit the model $E(C | V) = \beta V$$ to the data, with the
line constrained to pass through the origin." The constraint is
specified by the -1 term. 

We see that the estimate regression line is

$$E(C | V) = 0.98 V$$

## Full linear regression model

Say  we do not want to constrain the model to pass the line through the
origin. Our model is then

$$E(C | V) = \beta_0 + \beta_1 V$$

where we now have two unknown parameters to be estimated.

### Least-squares estimation, single predictor

Our sum of squared prediction errors is now

$$
\sum_{i=1}^{100} [O_{i} - (b_0 + b_1 V_{i}) ]^2
$$

This means setting two derivatives to 0 and solving. Since the
derivatives involve two different quantities to be optimized, $b_0$ 
and $b_1$, the derivatives are termed *partial*, and the $\partial$
symbol is used instead of 'd'.

$$
0 = 
\frac{\partial}{\partial b_0}\sum_{i=1}^{100} [O_{i} - (b_0 + b_1 V_{i}) ]^2
=
-2 \sum_{i=1}^{100} [O_{i} - (b_0 + b_1 V_{i}) ] 
$$

and

$$
0 = 
\frac{\partial}{\partial b_1}\sum_{i=1}^{100} [O_{i} - (b_0 + b_1 V_{i}) ]^2
=
-2 \sum_{i=1}^{100} [O_{i} - (b_0 + b_1 V_{i}) V_i] 
$$

We could then solve for the $b_i$, but let's go straight to the general
case.

## Least-squares estimation, general number of predictors

