
```{r}
#| include: false
library(dsld)
library(qeML)
```

{{< pagebreak >}}

 
# Shrinkage Estimators

::: {.callout-note}

## Goals of this chapter:

In the previous chapter, we introduced the norm of a vector. In many
data science applications, solutions with smaller norms may be more
accurate. This point is explored here.

:::

## Ridge Regression

In seminal paper, Hoerl and Kennard[*Technometrids*,
Februaru1970]{.column-margin} discussed the problem of
*multicollinearity* of predictor variables in a linear model.  

### Crux of the problem

Multicollinearity referred to settings in which the following concerns
arise:

* One column of the matrix $A$ in @eq-linregformula is nearly equal to
  some linear combination of the others.

* Thus $A$ is nearly not of full rank.

* Thus $A'A$ is nearly not of full rank.

* Thus $\widehat{\beta}$ is unstable, in the form of high variance.

Needless to say, the word "nearly" is vague, and leaves open the
question of "How near is 'near'?" and "What can we do about it?" Ridge
regression is Hoerl'ss and Kennards's answer to the latter question, and
it is based on a parameter that can be used to explore the former one.

### The ridge solution

Their solution is simple; Add some quantity to the diagonal of $A'A$.
Specifically, @eq-linregformula now becomes

$$
\widehat{\beta} - (A'A + \lambda I)^{-1} A'S
$${#eq-ridge}

where $\lambda$ is a positive number chosen by the analyst.

Here $A$ has dimensions $n \textrm{ x } p$, and $I$ is the $p \textrm{ x } p$
identity matrix.

### Matrix formulation

Using partitioned matrices helps understand ridge. Replace  $A$ and $S$
by

$$
A_{new} =
 \left (
 \begin{array}{r}
 A \\
 \lambda I \\
 \end{array}
 \right )
$$

and

$$
S_{new} =
 \left (
 \begin{array}{r}
 S \\
 0 \\
 \end{array}
 \right )
$$

where 0 means $p$ 0s. In essence, we are adding artificial data here,
consisting of $p$ new rows to $A$, and $p$ new elements to $S$.
So @eq-ridge is just the result of applying @eq-linregformula to
$A_{new}$ and $S_{new}$.

Loosely speaking, we can think of the addition of $\lambda I$ to $A'A$
makes the latter "larger", and thus its inverse smaller. In other words,
we are "shrinking" $\widehat{\beta}$ towards 0. This effect is made even
stronger by the fact that we added 0s data to $S$. This will be made
more precise below.

❄️  **Your Turn:** Show that $A_{new}$ is of full rank, $p$.

### Modern view

There are many ways to deal with multicollinearity, which we must remind
the reader only concerns *approximate* linear dependence pf the columns of
$A$. These days, ridge is often used for situations in which there is
*exact* linear dependence.

We saw such a setting in @sec-censusrref. There we deliberately induced
exact linear dependence by inclusion of both male and female dummy
variables. Let's apply ridge:

```{r}
library(qeML) 
data(svcensus) 
svc <- svcensus[,c(1,4:6)]  
svc$man <- as.numeric(svc$gender == 'male') 
svc$woman <- as.numeric(svc$gender == 'female') 
svc$gender <- NULL 
a <- svc[,-2] 
lambda <- 0.1 
a <- as.matrix(a) 
tmp1 <- solve(t(a) %*% a + lambda * diag(4)) 
tmp1 %*% t(a) %*% as.matrix(svc$wageinc) 
```

The results essentially are the same as what we obtained by having only
one dummy, thus no linear dependence: Men still enjoy about an \$11,000
advantage. But ridge allowed us to avoid deleting one of our dummies.
Such deletion is easy in this case, but for large $p$, say in the
hundreds or even more, some analysts prefer the convenience of ridge.

Of course, there is still the issue of how to choose the value of
$\lambda$. This is beyond the scope of this book, which focuses on linar
algebra.

### Formalizing the notion of shrinkage

In the early 1980s, the statistical world was shocked by research by
James and Stein that found, in short that:

<blockquote>

Say $W$ has $q$-dimensional normal distribution with mean vector $\mu$
and independent components having variance $\sigma^2$, each. We have
just $n = 1$ observation, $W$.  Then if $q \geq 3$, in terms of Mean
Squared Estimation Error, the best estimator of $\mu$ is NOT 
$W$. Instead, it's

$$
\left (
1 - \frac{(q-2) \sigma^2}{||W||^2} 
\right )
W
$$

</blockquote>

The quantity within the parentheses is typically smaller than 1, giving
us the shrinkage property.[E.g. if $\mu = 0$, then $E(||W||^2) = q
\sigma^2$]{.column-margin}

In the case of linear regression, shrinkage works there too, with $q$
being the number of columns in the $A$ matrix. 

### Shrinkage through length penalization

Say instead of minimizing @eq-matrixss, we minimize

$$
(S - Ab)'(S - Ab) + \lambda ||b||^2 
$${#eq-matrixssridge}

The larger $b$ is the harder it is to minimize the overall quantity
@eq-matrixssridge. We say that we *penalize* large values of $b$,
an indirect way of pursuing shrinkage. Now take the derivative and set
to 0:

$$
0 = A'(S-Ab) + \lambda b
$$

i.e.

$$
(A'A + \lambda I) b = A'S
$$ 

and thus

$$
\widehat{\beta} = (A'A+\lambda I)^{-1} A'S
$$

It's ridge! So here is formalization of our "almost singular" etc.
language above to justify ridge as a shrinkage estimator..

## The LASSO

The LASSO (Least Absolute Shrinkage and Selection Operator) was
developed by Robert Tibshirani in 1996, following earlier work by Leo
Breiman.  It takes $\widehat{\beta}$ to be the value of $b$ that
minimizes

$$
(S - Ab)'(S - Ab) + \lambda ||b||_1
$${#eq-matrixsslasso}

where the "l1 norm" is 

$$
||b|| = \sum_{i=1}^p |b_i|
$$

This is a seemingly minor change, but with important implications.  What
Breiman and Tibshirani were trying to do was to obtain a $\widehat{\beta}$
with lots of 0s, thereby providing a method for predictor variable
selection. This is important because so-called "parsimonious" prediction
models are desirable.

## A Warning

::: {.callout-important}

## No Statistical Inference on Shrinkage Estimators

Shrinkage produces a bias, of unknown size. Thus classical statistical
inference (confidence intervals, hypothesis tests), e.g. those based on
@eq-varcovar for linear models, is not possible.

:::



