
```{r}
#| include: false
library(dsld)
library(qeML)
```

{{< pagebreak >}}

 
# Shrinkage Estimators

::: {.callout-note}

## Goals of this chapter:

In the previous chapter, we introduced the norm of a vector. In many
data science applications, solutions with smaller norms may be more
accurate. This point is explored here.

:::

## Ridge Regression

In seminal paper, Hoerl and Kennard[*Technometrids*,
Februaru1970]{.column-margin} discussed the problem of
*multicollinearity* of predictor variables in a linear model.  

### Crux of the problem

Multicollinearity referred to settings in which the following concerns
arise:

* One column of the matrix $A$ in @eq-linregformula is nearly equal to
  some linear combination of the others.

* Thus $A$ is nearly not of full rank.

* Thus $A'A$ is nearly not of full rank.

* Thus $\widehat{\beta}$ is unstable, in the form of high variance.

Needless to say, the word "nearly" is vague, and leaves open the
question of "How near is 'near'?" and "What can we do about it?" Ridge
regression is Hoerl'ss and Kennards's answer to the latter question, and
it is based on a parameter that can be used to explore the former one.

### The ridge solution

Their solution is simple; Add some quantity to the diagonal of $A'A$.
Specifically, @eq-linregformula now becomes

$$
\widehat{\beta} - (A'A + \lambda I)^{-1} A'S
$${#eq-ridge}

where $\lambda$ is a positive number chosen by the analyst.

Here $A$ has dimensions $n \textrm{ x } p$, and $I$ is the $p \textrm{ x } p$
identity matrix.

### Matrix formulation

Using partitioned matrices helps understand ridge. Replace  $A$ and $S$
by

$$
A_{new} =
 \left (
 \begin{array}{r}
 A \\
 \lambda I \\
 \end{array}
 \right )
$$

and

$$
S_{new} =
 \left (
 \begin{array}{r}
 S \\
 0 \\
 \end{array}
 \right )
$$

where 0 means $p$ 0s. In essence, we are adding artificial data here,
consisting of $p$ new rows to $A$, and $p$ new elements to $S$.
So @eq-ridge is just the result of applying @eq-linregformula to
$A_{new}$ and $S_{new}$.

Loosely speaking, we can think of the addition of $\lambda I$ to $A'A$
makes the latter "larger", and thus its inverse smaller. In other words,
we are "shrinking" $\widehat{\beta}$ towards 0. This effect is made even
stronger by the fact that we added 0s data to $S$. This will be made
more precise below.

### Modern view

There are many ways to deal with multicollinearity, which we must remind
the reader only concerns *approximate* linear dependence pf the columns of
$A$. These days, ridge is often used for situations in which there is
*exact* linear dependence.

We saw such a setting in @sec-censusrref. There we deliberately induced
exact linear dependence by inclusion of both male and female dummy
variables. Let's apply ridge:

```{r}
library(qeML) 
data(svcensus) 
svc <- svcensus[,c(1,4:6)]  
svc$man <- as.numeric(svc$gender == 'male') 
svc$woman <- as.numeric(svc$gender == 'female') 
svc$gender <- NULL 
a <- svc[,-2] 
lambda <- 0.1 
a <- as.matrix(a) 
tmp1 <- solve(t(a) %*% a + lambda * diag(4)) 
tmp1 %*% t(a) %*% as.matrix(svc$wageinc) 
```

The results essentially are the same as what we obtained by having only
one dummy, thus no linear dependence: Men still enjoy about an \$11,000
advantage. But ridge allowed us to avoid deleting one of our dummies.
Such deletion is easy in this case, but for large $p$, say in the
hundreds or even more, some analysts prefer the convenience of ridge.

Of course, there is still the issue of how to choose the value of
$\lambda$. This is beyond the scope of this book, which focuses on linar
algebra.

### Formalizing the notion of shrinkage

JAMES-STEIN

EQUIV TO RIDGE; MATRIX DERIVS; THEN LASSO





