
```{r} 
#| include: false
library(dsld)
library(qeML)
```

{{< pagebreak >}}

# Singular Value Decomposition

::: {.callout-note}

## Goals of this chapter:

We've seen the notion of eigenanalysis for square matrices. Well, it
turns out this can be extended usefully for nonsquare matrices, via
*Singular Value Decomposition* (SVD), the topic of this chapter.

::: 



```{r} 
#| include: false
library(dsld)
library(qeML)
```

## Basic Idea

We'll take this as our definition of SVD:

<blockquote>

Given a (very possibly nonsquare) matrix $A$, we wish to find orthogonal
matrices $U$ and $V$, and a diagonal matrix $\Sigma$ whose nonzero
elements are positive, such that

$$
A = U \Sigma V'
$$

</blockquote>

## Solution

1. Compute the eigenvalues and eigenvectors of $A'A$. 

2. Since $A'A$ is symmetric and positive-semidefinite, its eigenvalues 
   will be nonnegative and its length-1 eigenvectors will be orthogonal. 

3. Set $\Sigma$ to $diag(\sqrt{\sigma_1},...,\sqrt{\sigma_r}$, the
   nonincreasing list of postive eigenvalues. Set the columns of $V$ to
   the corresponding eigenvectors.

4.  Set $U$ as with $V$, but based on $AA'$ rather than $$A'A$.$

